\section{Introduction}
\label{sec:introduction}

Network security remains a central challenge for modern computing infrastructure. Traditional intrusion detection systems (IDS) rely on signatures or static statistical rules, which are inherently brittle against zero-day attacks, distribution shift, and adversarial evasion \cite{liao2024survey, moustafa2023explainable, wang2023robustnessnids}. Recent learning-based IDS research has pushed toward self-supervised generalization \cite{golchin2024ssclids, xu2024gnnssl}, open-set recognition \cite{qiu2024vaemax}, and transformer-based feature modeling \cite{xi2024idsmtran, xin2025crossdatasettransformer}. Yet a deeper problem remains largely unaddressed: most IDS training pipelines assume a \emph{passive} threat environment, where the attacker does not adapt to the deployed model. Empirical evidence shows that classifiers with strong in-distribution benchmark scores can degrade sharply under cross-dataset testing or targeted perturbations \cite{cantone2024crossdataset, wang2023robustnessnids}, precisely because they are not trained against a strategically adaptive adversary.

Reinforcement learning (RL) offers a natural framework for adaptive defense, yet existing RL-based IDS pipelines overwhelmingly train the defender against either static disturbances or a \emph{simultaneously} updated attacker \cite{malialis2015multiagent, anderson2016rl_security}. Simultaneous updates produce co-adaptation rather than genuine worst-case hardening: neither agent is explicitly optimizing against the other's best response. This distinction matters. A Stackelberg game formulation---where the defender commits to a strategy and the attacker best-responds---provides a stronger robustness objective than ordinary multi-agent co-training \cite{vonStackelberg1952, colson2007bilevel, pinto2017robust_rl}.

This paper makes the Stackelberg formulation operational. We propose \textbf{Bi-level Adversarial Reinforcement Learning (Bi-ARL)}, in which the outer defender update is conditioned on an inner attacker that is iterated to approximate a best response before each outer step. We then recognize that the PPO backbone used in RL is a relatively weak tabular detector, and transfer the same bilevel adversarial training principle to modern supervised detectors under the \textbf{BiAT (Bilevel Adversarial Training)} framework, yielding \textbf{BiAT-MLP} and \textbf{BiAT-FTTransformer} \cite{gorishniy2021fttransformer}.

\subsection{Contributions}

\begin{enumerate}
    \item \textbf{Bi-ARL framework.} We formalize IDS training as a bi-level Stackelberg optimization problem and implement a practical PPO-based pipeline for tabular network-flow benchmarks, including a convergence criterion based on KL divergence of the attacker policy.

    \item \textbf{BiAT framework: bilevel adversarial training for supervised detectors.} We transfer the core bilevel training idea to supervised tabular detectors under the BiAT (Bilevel Adversarial Training) framework, producing BiAT-MLP and BiAT-FTTransformer. This preserves the Stackelberg training principle while exploiting stronger tabular backbones.

    \item \textbf{Theoretical analysis of finite-step bilevel training.} We derive approximation error bounds for the finite-step inner loop (Propositions~\ref{prop:inner_approx} and~\ref{prop:defender_subopt}), quantifying the trade-off between inner-loop budget $K$ and Stackelberg equilibrium quality.

    \item \textbf{Inner-loop stabilization.} We identify and address the inner-loop instability problem through entropy-regularized Stackelberg training with exponential decay scheduling and an attacker warm-up schedule, supported by theoretical motivation from regularized MDPs \cite{geist2019entropy}.

    \item \textbf{Strongest neural performance on UNSW-NB15.} BiAT-FTTransformer achieves 89.51\% mean F1 and 14.97\% mean FPR on UNSW-NB15, the lowest false positive rate among all neural detectors evaluated, trailing LightGBM by only 2.35 F1 points.

    \item \textbf{Systematic adversarial robustness study.} We evaluate FGSM robustness across all model families and provide PGD evaluation and statistical significance testing, enabling an honest characterization of the framework's capabilities and limitations.
\end{enumerate}
