\section{Related Work}
\label{sec:related_work}

\subsection{Intrusion Detection Systems}

Traditional IDSs rely on signature matching, which remains effective for known attack templates but fails on zero-day threats or rapidly evolving adversaries. Surveys by Liao \emph{et al.} \cite{liao2024survey} and Moustafa \emph{et al.} \cite{moustafa2023explainable} document the breadth of learning-based IDS approaches, spanning classical machine learning, deep neural networks, graph-based methods, and IoT-oriented pipelines. Beyond accuracy, a growing body of work highlights that many public NIDS benchmarks contain outdated traffic patterns, shortcut features, or overly permissive split assumptions that inflate deployment-time estimates \cite{datasetreview2025, ring2019nids}. This concern is directly relevant here: the empirical behavior of Bi-ARL changes substantially across NSL-KDD \cite{tavallaee2009nslkdd}, UNSW-NB15 \cite{moustafa2015unswnb15}, and CIC-IDS2017 \cite{sharafaldin2018cicids2017}, echoing the cross-dataset instability documented by Cantone \emph{et al.} \cite{cantone2024crossdataset}.

\subsection{Generalization and Modern Deep IDS}

An important recent trend is to move beyond single-dataset, closed-set evaluation. Golchin \emph{et al.} proposed SSCL-IDS \cite{golchin2024ssclids}, a self-supervised contrastive approach that improves detection under unseen traffic distributions. Xu \emph{et al.} extended self-supervised learning to graph neural network flow modeling, showing competitive performance with supervised methods while reducing label dependence \cite{xu2024gnnssl}. On the architecture side, IDS-MTran \cite{xi2024idsmtran} applies multi-scale transformer branches to NSL-KDD and UNSW-NB15, and the cross-dataset transformer study by Xin and Xu \cite{xin2025crossdatasettransformer} adds calibration-oriented analysis across the same benchmarks. The FT-Transformer of Gorishniy \emph{et al.} \cite{gorishniy2021fttransformer}, originally proposed for general tabular learning, provides the backbone for our BiAT-FTTransformer. These works collectively define a modern deep-learning comparison set that goes well beyond legacy RL-only baselines---but they still optimize static prediction models and do not incorporate an adaptive attacker into the training loop. The BiAT framework proposed in this paper directly targets this gap.

\subsection{Robustness, Open-Set Detection, and Cross-Dataset Evaluation}

Wang \emph{et al.} \cite{wang2023robustnessnids} argue that adversarial perturbations and distribution shift should be treated as first-class evaluation criteria for ML-based NIDSs. Apruzzese \emph{et al.} \cite{apruzzese2022adversarial_nids} systematically model realistic adversarial attacks against network classifiers, showing significant degradation under crafted perturbations. Cantone \emph{et al.} \cite{cantone2024crossdataset} provide direct evidence that within-dataset performance can deteriorate sharply under cross-dataset evaluation. In the open-set direction, VAEMax \cite{qiu2024vaemax} combines OpenMax with a variational autoencoder to handle previously unseen attack classes. Newer dataset initiatives---SHIELD/NF-UQ-NIDS \cite{shield2025, sarhan2021nfuqnids} and 5G-NIDD \cite{siriwardhana2025nidd}---indicate the field is moving toward more standardized, contemporary flow collections. The adversarial robustness analysis in this paper extends this direction by applying FGSM \cite{goodfellow2015fgsm} and PGD \cite{madry2018pgd} evaluations to both RL-based and supervised-bilevel models.

\subsection{Adversarial Machine Learning Foundations}

The threat of adversarial examples---small, crafted input perturbations that cause misclassification---was formalized by Goodfellow \emph{et al.} \cite{goodfellow2015fgsm} and extended to iterative PGD attacks by Madry \emph{et al.} \cite{madry2018pgd}. Carlini and Wagner \cite{carlini2017cw} demonstrated that many defenses certified against FGSM still fail under stronger optimization-based attacks. Zhang \emph{et al.} \cite{zhang2019trades} characterized the fundamental robustness-accuracy trade-off via TRADES, and Biggio \emph{et al.} \cite{biggio2013evasion} established the evasion attack paradigm in the security domain. Yuan \emph{et al.} \cite{yuan2019adv_survey} provide a comprehensive survey. Our BiAT framework models use PGD-style inner loops during training, directly inheriting the adversarial training framework of Madry \emph{et al.}

\subsection{Adversarial Reinforcement Learning and Stackelberg Games}

RL applied to security games naturally accommodates sequential attacker-defender interaction \cite{malialis2015multiagent, anderson2016rl_security}. Standard multi-agent RL solutions---including Nash-Q \cite{hu2003nashq} and MADDPG \cite{lowe2017maddpg}---update both agents simultaneously, which can result in co-adaptation rather than a genuine worst-case adversary. The Stackelberg formulation \cite{vonStackelberg1952, colson2007bilevel} addresses this by designating one player as the leader who commits to a strategy before the follower responds. Bilevel optimization \cite{sinha2018bilevel, bard1998bilevel} provides the mathematical machinery to realize this: the outer problem optimizes the leader against an inner problem that computes the follower's best response. Pinto \emph{et al.} \cite{pinto2017robust_rl} apply a similar structure to robust RL in physical control. Li \emph{et al.} \cite{li2022advrl_security} extend adversarial RL to software-defined networks. Our Bi-ARL explicitly implements the bilevel Stackelberg structure with a convergence-tested inner loop, distinguishing it from simultaneous-update MARL baselines. Empirically, we also show the limits of this approach on modern datasets, motivating the transfer to stronger supervised backbones under the BiAT framework.

\subsection{Adversarial Robustness in Tabular Domains}

Adversarial training in tabular settings differs from image domains: feature semantics and constraints vary by column, and perturbations must respect domain validity \cite{apruzzese2022adversarial_nids, biggio2013evasion}. The TRADES framework \cite{zhang2019trades} provides a principled objective for balancing clean accuracy with adversarial robustness, and the BiAT framework loss $L = (1-\lambda)L_{\text{clean}} + \lambda L_{\text{adv}}$ is a direct instantiation of this trade-off. Unlike image adversarial training, the tabular context requires careful tuning of perturbation magnitude and categorical-feature exclusion, both of which are addressed in our experimental configuration.
