\section{Conclusion}
\label{sec:conclusion}

This paper began with Bi-level Adversarial Reinforcement Learning (Bi-ARL), which formalizes attacker-defender interaction as a Stackelberg game and trains the detector against an adversary that is repeatedly adapted toward its best response. The core bilevel training principle was then transferred to supervised tabular detectors under the \textbf{BiAT (Bilevel Adversarial Training)} framework, yielding BiAT-MLP and BiAT-FTTransformer.

On NSL-KDD, Bi-ARL achieves the best mean F1 among RL baselines (80.17\%) and reduces FPR by 29.16 percentage points versus vanilla PPO---the primary measurable benefit of Stackelberg bilevel training on the controlled benchmark. On UNSW-NB15, BiAT-FTTransformer reaches 89.51\% mean F1 and 14.97\% mean FPR, the lowest false positive rate among all neural detectors evaluated, while trailing LightGBM by only 2.35 F1 points. FGSM robustness analysis reveals partial robustness at small perturbation budgets, with catastrophic failure at $\epsilon=0.3$ identifying inner-loop convergence quality as the main limitation. The current contribution is therefore best characterized as a \textit{competitive bilevel adversarial training framework with a strong tabular backbone}, not a universally superior IDS system.

The experimental infrastructure developed alongside this paper---extended robustness evaluation (FGSM + PGD across all model families), a Vanilla FT-Transformer clean-only baseline for honest attribution, hyperparameter sensitivity analysis ($K_{\text{inner}}$ and $\lambda$), and Wilcoxon signed-rank significance testing with Cohen's $d$ effect sizes---provides a solid empirical foundation. The most promising future directions are: (1) stabilizing the inner loop on modern datasets through entropy regularization or attacker warm-up schedules; (2) evaluating on CSE-CIC-IDS2018 \cite{sharafaldin2018cicids2018} and NF-UQ-NIDS \cite{sarhan2021nfuqnids} to validate generalization beyond the current three benchmarks; and (3) exploring the combination of bilevel adversarial training with boosted-tree ensemble methods to close the remaining F1 gap against LightGBM.
