
\documentclass[journal]{IEEEtran}

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{multirow}  % 支持表格多行合并

% 为算法环境定义缺失的命令
\algrenewcommand\algorithmicrequire{\textbf{Input:}}
\algrenewcommand\algorithmicensure{\textbf{Output:}}
\renewcommand{\Comment}[1]{\State \(\triangleright\) #1}
% 定义大写命令别名以兼容现有代码
\let\STATE\State
\let\FOR\For
\let\ENDFOR\EndFor
\let\IF\If
\let\ENDIF\EndIf
\let\WHILE\While
\let\ENDWHILE\EndWhile

\begin{document}

\title{Bi-level Adversarial Reinforcement Learning for Robust Network Threat Detection}

\author{Jun Zhu
    \thanks{J. Zhu is with the School of Cyber Science and Engineering, Nanjing University of Science and Technology, Nanjing, China.}
}

\maketitle

\begin{abstract}
    Network threat detection faces significant challenges from adaptive attackers, dataset shift, and evasion-oriented perturbations. This paper starts from a \textbf{Bi-level Adversarial Reinforcement Learning (Bi-ARL)} formulation and then refactors the same bilevel idea into a stronger supervised tabular detector. We evaluate both the original RL line and the refactored route-C models on three public intrusion-detection benchmarks: NSL-KDD, UNSW-NB15, and CIC-IDS2017. On NSL-KDD, Bi-ARL still achieves the best mean F1 score among the RL variants (\textbf{80.17\%}) and substantially reduces false positives relative to vanilla PPO (10.41\% vs.\ 39.57\%). On the more modern UNSW-NB15 benchmark, the strongest route-C model, \textbf{BiAT-FTTransformer}, reaches \textbf{89.51\%} mean F1 and \textbf{14.97\%} mean FPR, making it the strongest neural detector in our study and yielding the lowest false positive rate among all compared methods on that dataset. On CIC-IDS2017 under a random stratified split, however, strong tree-based tabular baselines still dominate with about \textbf{99.7\%} mean F1, while route-C remains clearly behind. These findings support a narrower but stronger conclusion: explicit bilevel adversarial training is useful, the FT-Transformer backbone materially improves the modern-data results, but the overall framework is not yet state of the art against the strongest boosted-tree baselines.
\end{abstract}

\begin{IEEEkeywords}
    Network Threat Detection, Adversarial Reinforcement Learning, Bi-level Optimization, Intrusion Detection Systems, Game Theory.
\end{IEEEkeywords}

\input{sections/01_introduction}
\input{sections/02_related_work}
\input{sections/03_methodology}
\input{sections/04_experiments}
\input{sections/05_conclusion}

% Bibliography
\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
